Reduced plasticity and relative variance in physiological rates of ectotherm populations under climate change

### 2024-05-17 ms.qmd
Revising intro

---
title: "Reduced plasticity and variance in physiological rates of ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
format:
  docx:
    reference-doc: ./bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining phenotypic variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. Using newly developed effect size estimates and meta-analysis (>1900 effects from 226 species), we show that across habitats relative variance in physiological rates decreases at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that beneficial acclimation responses in some ecosystems (marine and freshwater) may trade-off with reductions in physiological rate variance that can have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

Climate change is expected to result in warmer but also more variable thermal environments globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past should result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one].  Without plasticity or adaptation, high extinction rates are predicted unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates, thereby buying time for genetic adaptation to take place (i.e., the 'plasticity first hypothesis')[@WestEberhard2003; @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant (but see [@Leung2020]), likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], or the prediction is only supported for specific life-history stages.

Reversible phenotypic plasticity, such as acclimation, is expected to provide greater potential to buffer populations from climate impacts as responses are relatively rapid and can therefore be fine-tuned to proximate environmental conditions (assuming the costs of plasticity are low) [@Dewitt1998; @Scheiner2020]. Acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to facilitate a rapid response to the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. However, the focus up to now has been primarily on mean physiological responses. For example, mean thermal tolerances or acclimation capacities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022]. However, it is possible that intrapopulation variability might also be impacted in addition to the mean. Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change (i.e., the portfolio effect) [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013; @Sanderson2023]. Decreases in phenotypic variance also suggests strong stabilising selection or reflects constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation; @Sanderson2023]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date there are few data testing the importance of variances in this context [@Scheiner2020].

@@

Where, $R_{1}$ and  $R_{2}$ are mean physiological rates and  $T_{1}$ and  $T_{2}$ are the temperatures at which these rates are measured. Log transformation of this ratio makes the effect size normally distributed. @eq-lnq10 is essentially a temperature corrected equivalent to the log response ratio (lnRR) [@Hedges1999-de; @Lajeunesse2011-fq] when the numerator and denominator are measured at different temperatures. This allows comparisons of the means from two temperature treatments directly regardless of the absolute measurement temperatures. The sampling variance for @eq-lnq10 can be computed as follows (as described in @Noble2022-ty):

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
Using the mean, standard deviation, and sample size for all acute and acclimation treatments of studies in our databases we derived acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For all effect sizes the higher acute or acclimation temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean or variance is larger at the higher of the two temperatures, standardized to 10&deg;C.

@@

## Data and code availability 
All data and code used to reproduce analyses can be found on GitHub at: https://github.com/daniel1noble/Q10_meta_analysis.

>>>>

---
title: "Reduced plasticity and relative variance in physiological rates of ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
format:
  docx:
    reference-doc: ./bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining phenotypic variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. We developed effect size estimates that quantify how physiological variance changes relative to the mean as temperature increases, which we refer to as 'relative variance'. We show by meta-analysing >1900 effects from 226 species that relative variance in physiological rates decreases at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that beneficial acclimation responses in some ecosystems (marine and freshwater) may trade-off with reductions in physiological rate variance that can have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

Climate change is expected to result in warmer but also more variable thermal environments globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is predicted to result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one]. Without plasticity or adaptation, high extinction rates are expected unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016].

Phenotypic plasticity is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to proximate environmental conditions making it the first ‘line-of-defense’ against rapid environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses. Acute temperature increases that push organisms from their thermal optimum can be mitigated by plastically adjusting reaction rates in response to chronic temperatures through acclimation [@Seebacher2015-wi; @Havird2020-mj]. Acclimation to higher temperatures does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather rates remain similar to acute responses at high temperatures ('no acclimation') or only partially compensate [@huey1999testing; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. However, the focus up to now has been primarily on mean physiological responses neglecting how variability in physiological rates might also be impacted at higher temperatures.

As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability might also be impacted. NEED MORE ABOUT WHY VARIANCE NOT JUST IMPORTANCE

Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change (i.e., the portfolio effect) [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013; @Sanderson2023]. Decreases in phenotypic variance also suggests strong stabilising selection or reflects constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation; @Sanderson2023]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date there are few data testing the importance of variances in this context [@Scheiner2020].

@@

Where, $R_{1}$ and  $R_{2}$ are mean physiological rates and  $T_{1}$ and  $T_{2}$ are the temperatures at which these rates are measured. $R_{1}$ and $R_{2}$ can be taken from samples of organisms measured acutely at two temperatures or after having been acclimated at these same temperatures. Log transformation of this ratio makes the effect size normally distributed. @eq-lnq10 is essentially a temperature corrected equivalent to the log response ratio (lnRR) [@Hedges1999-de; @Lajeunesse2011-fq] when the numerator and denominator are measured at different temperatures. This allows comparisons of the means from two temperature treatments directly regardless of the absolute measurement temperatures. The sampling variance for @eq-lnq10 can be computed as follows (as described in @Noble2022-ty):

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
As indicated above, $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures. For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean or variance is larger at the higher of the two temperatures, standardized to 10&deg;C.

@@

## Data and code availability 
All data and code used to reproduce analyses can be found on GitHub at: https://github.com/daniel1noble/Q10_meta_analysis and is deposited in Zenodo (https://doi.org/10.5281/zenodo.11123600).
